---
title: "ggplotly and plotly keyboard accessibility"
format: html
vignette: >
  %\VignetteIndexEntry{ggplotly-and-plotly-keyboard-accessibility}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r, setup}
library(ROvis.plotly) # nolint: missing_package_linter.
```
# Introduction
This vignette demonstrates the use of accessibility functions for interactive `plotly` and `ggplotly` plots.
These functions are designed to improve keyboard and screen reader accessibility for various types of interactive charts.

Note: The example plots in this vignette are for demonstration only and do not themselves comply with the Rijksoverheid style requirements.

Key Accessibility Features

- Keyboard navigation: Users can focus bars, points, and legend items using Tab and Arrow keys.
- Accessible tooltips: Tooltip content is made available for screen readers and keyboard navigation.
- Screen reader support: Meaningful ARIA labels are provided for plot elements.
- Legend accessibility: Legends can be reached and toggled via keyboard.
- Toolbar control: The plotly modebar (toolbar) can be disabled for improved accessibility.

The functions covered in this vignette are:

- ro_ply_keyboard_nav_barchart()
- ro_ply_keyboard_nav_bargraph()
- ro_ply_keyboard_nav_trendline()
- ro_ply_keyboard_nav_linechart()
- ro_ply_keyboard_nav_legend()
- ro_ply_disable_toolbar()


# Keyboard accessibility - ggplotly and plotly plots
Each section below demonstrates how to add accessibility to a ggplotly or plotly graph using the functions from this package.
For each plot type, you will see an example with and without accessibility enhancements.

## Horizontal Barchart
This section demonstrates how to make a horizontal bar chart created with ggplotly or plotly keyboard accessible. The example shows how to use `ro_ply_keyboard_nav_barchart()` to enable keyboard navigation and accessible tooltips for all bars.

### Example data
```{r example data horizontal barchart}
data <- tidyr::tibble(
  expand.grid(`Age group` = c("A", "B", "C", "D", "E"), `Sex` = c("Man", "Vrouw")),
  `Number of Cases` = sample(1:50, 10, TRUE)
) 
```

### ggplotly - Horizontal Barchart
```{r ggplotly horizontal barchart}
library(ggplot2)
library(plotly)
 
p <- ggplot(data, aes(
  x = `Number of Cases`,
  y = `Age group`,
  fill = Sex,
  text = paste0(
    "Leeftijdsgroep: <b>", `Age group`, "</b><br>",
    "Geslacht: <b>", Sex, "</b><br>",
    "Aantal cases: <b>", `Number of Cases`, "</b>"
  )
)) +
  geom_col(position = "dodge") +
  labs(
    x = "Aantal cases",
    y = "Leeftijdsgroep",
    fill = "Geslacht"
  ) 

# Convert to plotly and add accessibility
fig <- ggplotly(p, tooltip = "text")

# Add the keyboard accessibility in the plot
fig <- ro_ply_keyboard_nav_barchart(fig)
fig
```

### plotly - Horizontal Barchart
```{r plotly horizontal barchart}
library(plotly)
 
fig <- plot_ly(data = data,
  x = ~`Number of Cases`,              
  y = ~`Age group`,                    
  color = ~`Sex`,          
  type = "bar",
  orientation = "h",                   # horizontal bars
  hovertext  = ~paste0(
    "Leeftijdsgroep: <b>", `Age group`, "</b><br>",
    "Geslacht: <b>", `Sex`, "</b><br>",
    "Aantal cases: <b>", `Number of Cases`, "</b>"
  ),
  text  = ~paste0(
    "Leeftijdsgroep: <b>", `Age group`, "</b><br>",
    "Geslacht: <b>", `Sex`, "</b><br>",
    "Aantal cases: <b>", `Number of Cases`, "</b>"
  ),    
  textposition = "none",
  hoverinfo = "text"
) 
# Add the keyboard accessibility in the plot
fig <- ro_ply_keyboard_nav_barchart(fig)
fig
 
```
## Grouped Bar Graph / Column Chart
This section demonstrates how to make a column chart created with ggplotly or plotly keyboard accessible. The example shows how to use `ro_ply_keyboard_nav_barchart()` to enable keyboard navigation and accessible tooltips for all bars.

### Example data
```{r example data column chart}
age_levels <- c(
  "0", "1-9", "10-19", "20-29",
  "30-39", "40-49", "50-59",
  "60-69", "70-79", "80-89", "90+"
)

base_data <- tidyr::expand_grid(
  `Age group` = age_levels,
  Sex = c("Men", "Women")
) |>
  dplyr::mutate(
    `Number of Cases` = sample(1:50, length(age_levels) * 2, TRUE)
  )

summary_data <- base_data |>
  dplyr::summarize(
    `Number of Cases` = sum(`Number of Cases`),
    .by = `Age group`
  ) |>
  dplyr::mutate(Sex = "Total")

data <- dplyr::bind_rows(base_data, summary_data) |>
  dplyr::mutate(
    `Age group` = factor(`Age group`, levels = age_levels),
    Sex = factor(Sex, levels = c("Men", "Women", "Total"))
  ) 

```
### ggplotly Grouped Bar Graph / Column Chart
```{r ggplotly column chart}
p <- ggplot(data, aes(
  x = `Age group`,
  y = `Number of Cases`,
  fill = Sex,
  text = paste0(
    "Geslacht: <b>", Sex, "</b>",
    "<br>Leeftijdsgroep: <b>", `Age group`,
    "</b><br>Aantal cases: <b>",
    format(round(`Number of Cases`, 0), big.mark = ".", decimal.mark = ",", scientific = FALSE), "</b>"
  )
)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.7) +
  labs(x = "Leeftijdsgroep", y = "Aantal cases", fill = "Geslacht")

# Convert to interactive plotly plot
fig <- ggplotly(p, tooltip = "text")
fig <- ro_ply_keyboard_nav_bargraph(fig)
fig
```
### plotly Grouped Bar Graph / Column Chart
```{r plotly column chart}
fig <- plot_ly(
  data = data,
  x = ~`Age group`,
  y = ~`Number of Cases`,
  type = "bar",
  color = ~Sex,
  text = ~paste0(
    "Geslacht: <b>", Sex, "</b>",
    "<br>Leeftijdsgroep: <b>", `Age group`,
    "</b><br>Aantal cases: <b>", 
    format(round(`Number of Cases`, 0), big.mark = ".", decimal.mark = ",", scientific = FALSE), "</b>"
  ),
  textposition = "none",
  hoverinfo = "text"
) |>
  layout(barmode = "group")

# Add the keyboard accessibility in the plot
fig <- ro_ply_keyboard_nav_bargraph(fig)
fig
```

## Bar Graph
This section demonstrates how to make a bar graph created with ggplotly or plotly keyboard accessible. The example shows how to use `ro_ply_keyboard_nav_barchart()` to enable keyboard navigation and accessible tooltips for all bars.

### Example data
```{r example data bargrpaph}
data <- tidyr::tibble(
  Sex = factor(c("Vrouw", "Man"), levels = c("Vrouw", "Man")),
  n = sample(1e6:5e6, 2)
)

```
### ggplotly Bar Graph
library(ggplot2)
library(plotly)
 
```{r ggplotly bar graph}
p <- ggplot(data, aes(x = Sex, y = n, fill = Sex,
                      text = paste0(
                        "Geslacht: <b>", Sex, "</b>",
                        "</b><br>Aantal cases: <b>",
                        format(round(n, 0), big.mark = ".", decimal.mark = ",", scientific = FALSE), "</b>"
                      ))) +
  geom_col(width = 0.6) +
  labs(x = "Geslacht", y = "Aantal cases") +
  theme_minimal()

# Convert to plotly and add accessibility
fig <- ggplotly(p, tooltip = "text")

# Add the keyboard accessibility in the plot
fig <- ro_ply_keyboard_nav_bargraph(fig)
fig
```
### plotly Bar Graph
```{r plotly bar graph}
fig <- plot_ly(
  data = data,
  x = ~Sex,
  y = ~n,
  type = "bar",
  color = ~Sex,
  text = ~paste0(
    "Geslacht: <b>", Sex, "</b>",
    "</b><br>Aantal cases: <b>", 
    format(round(n, 0), big.mark = ".", decimal.mark = ",", scientific = FALSE), "</b>"
  ),
  hoverinfo = "text",
  width = 0.6
) 

# Add the keyboard accessibility in the plot
fig <- ro_ply_keyboard_nav_bargraph(fig)
fig
```
## Stacked Bar Graph
This section demonstrates how to make a stacked bar graph created with ggplotly or plotly keyboard accessible. The example shows how to use `ro_ply_keyboard_nav_barchart()` to enable keyboard navigation and accessible tooltips for all bars.
 
### Example data
```{r example data stacked bar graph}
agegroups <- c("0-9", "10-19", "20-29", "30-39", "40-49", "50-59", "60-69", "70-79", "80-89", "90+")
sex <- c("Vrouw", "Man", "Onbekend")

data <- expand.grid(
  Agegroup = agegroups,
  Sex = sex,
  stringsAsFactors = FALSE
) |>
  tibble::as_tibble() |>
  dplyr::rowwise() |>
  dplyr::mutate(
    n = if (Sex == "Onbekend") sample(200:3000, 1) else sample(200000:700000, 1)
  ) |>
  dplyr::ungroup() |>
  dplyr::mutate(Sex = factor(Sex, levels = c("Vrouw", "Man", "Onbekend")))

```
### ggplotly Stacked Bar Graph
```{r ggplotly stacked bar graph}
library(ggplot2)
library(plotly)
 
p <- ggplot(data, aes(x = Agegroup, y = n, fill = Sex,
                      text = paste0(
                        "Aantal cases: <b>", format(n, big.mark = ".", decimal.mark = ",", scientific = FALSE),
                        "</b><br>Leeftijdsgroep: <b>", Agegroup,
                        "</b><br>Geslacht: <b>", Sex
                      ))) +
  geom_col(width = 0.7, position = "stack") +
  labs(x = "Leeftijdsgroep", y = "Aantal cases", fill = "Geslacht")

# Convert to plotly and add accessibility
fig <- ggplotly(p, tooltip = "text")

# Add the keyboard accessibility in the plot
fig <- ro_ply_keyboard_nav_bargraph(fig)
fig
```
### plotly Stacked Bar Graph
```{r plotly stacked bar graph}
library(plotly)
 
fig <- plot_ly(
  data = data, # your tibble
  x = ~Agegroup,
  y = ~n,
  color = ~Sex,
  type = "bar",
  text = ~paste0(
    "Aantal cases: <b>", format(n, big.mark = ".", decimal.mark = ",", scientific = FALSE),
    "</b><br>Leeftijdsgroep: <b>", Agegroup,
    "</b><br>Geslacht: <b>", Sex
  ),
  textposition = "none",
  hoverinfo = "text"
) |> 
  layout(barmode = "stack")

# Add the keyboard accessibility in the plot
fig <- ro_ply_keyboard_nav_bargraph(fig)
fig
```

## Proportional Stacked Bar Graph
This section demonstrates how to make a proportional stacked bar graph created with plotly keyboard accessible. The example shows how to use `ro_ply_keyboard_nav_barchart()` to enable keyboard navigation and accessible tooltips for all bars.

Note: Proportional stacked bar graphs are not supported via ggplotly, only via plotly.


### Example data
```{r example data proportional stacked bar graph}
age_cat <- c("0-19", "20-39", "40-59", "60+")
sex <- c("Vrouw", "Man")
period <- c("Na", "Voor")

data <- expand.grid(
  age_cat = age_cat,
  omicron_period = period,
  stringsAsFactors = FALSE
) |>
  tibble::as_tibble() |>
  dplyr::rowwise() |>
  dplyr::mutate(
    # Simuleer totaal aantal cases per groep
    total_n = sample(50000:1200000, 1),
    # Simuleer random verhouding tussen vrouw/man
    vrouw_prop = runif(1, 0.45, 0.55),
    man_prop = 1 - vrouw_prop,
    n_vrouw = as.integer(total_n * vrouw_prop),
    n_man = total_n - n_vrouw
  ) |>
  dplyr::ungroup() |>
  # Zet weer om naar long format per sex
  tidyr::pivot_longer(
    cols = c(n_vrouw, n_man),
    names_to = "Sex",
    values_to = "n",
    names_prefix = "n_"
  ) |>
  dplyr::group_by(age_cat, omicron_period) |>
  dplyr::mutate(
    prop = n / sum(n),
    bar_label = paste0(age_cat, " (", omicron_period, ")"),
    prop_label = paste0(format(round(prop * 100, 1), decimal.mark = ","), "%")
  ) |>
  dplyr::ungroup() |>
  dplyr::select(age_cat, Sex, omicron_period, n, prop, bar_label, prop_label)

# Order the labels
data <- data |>
  dplyr::mutate(
    Sex = dplyr::recode(Sex, "vrouw" = "Vrouw", "man" = "Man")
  )
bar_label_levels <- as.vector(outer(age_cat, period, function(a, p) paste0(a, " (", p, ")")))
data <- data |>
  dplyr::mutate(
    bar_label = factor(bar_label, levels = bar_label_levels),
    Sex = factor(Sex, levels = c("Vrouw", "Man"))
  ) |>
  dplyr::arrange(bar_label, Sex) |>
  dplyr::select(age_cat, Sex, omicron_period, n, prop, bar_label, prop_label)

# Show the structure of the data
str(data)
```

```{r ggplotly proportional stacked bar graph, include = FALSE}
### ggplotly Proportional Stacked Bar Graph
library(ggplot2)
library(plotly)
 
```
### plotly Proportional Stacked Bar Graph
```{r plotly proportional stacked bar graph}
library(plotly)
 
fig <- plot_ly(
  data = data,
  y = ~bar_label,
  x = ~prop,
  color = ~Sex,
  orientation = "h",
  type = "bar",
  text = ~paste0(
    "Leeftijdsgroep: <b>", age_cat, "<br></b>",
    "Omicron periode: <b>", omicron_period, "<br></b>",
    "Geslacht: <b>", Sex, "<br></b>",
    "Percentage: <b>", prop_label, "<br></b>"
  ),
  textposition = "auto",
  hovertemplate = ~paste0(
    "Leeftijdsgroep: <b>", age_cat, "<br></b>",
    "Omicron periode: <b>", omicron_period, "<br></b>",
    "Geslacht: <b>", Sex, "<br></b>",
    "Percentage: <b>", prop_label, "<br></b>"
  ),
  hoverinfo = "text"
) |> layout(barmode = "stack")

fig <- ro_ply_keyboard_nav_barchart(fig)
fig
```




## Trendline
This section demonstrates how to make a trendline created with ggplotly or plotly keyboard accessible. The `ro_ply_keyboard_nav_trendline()` function allows users to focus on individual points (not lines), with tooltips and ARIA labels for screen readers.

### Example data
```{r example data Trendline}
dates <- seq(as.Date("2021-10-04"), by = "day", length.out = 100)
n <- round(2500 + 700 * sin(seq_along(dates) / 25) + seq_along(dates) * 3 + rnorm(length(dates), 0, 150))
moving_average <- stats::filter(n, rep(1 / 7, 7), sides = 2)
text <- paste0(
  "Dag: <b>", format(dates, "%d-%m-%Y"), "</b><br>",
  "Aantal cases: <b>", format(n, big.mark = ".", decimal.mark = ",", scientific = FALSE), "</b><br>",
  "7-daags gemiddelde: <b>", format(round(moving_average, 0), 
                                    big.mark = ".", decimal.mark = ",", scientific = FALSE), "</b>"
)

data <- tibble::tibble(
  Date_statistics = dates,
  n = n,
  moving_average = as.numeric(moving_average),
  text = text
)
```
### ggplotly Trendline
```{r ggplotly Trendline}
p <- ggplot(data, aes(x = Date_statistics, y = n, text = text, group = 1)) +
  geom_line() +
  geom_point() 

# Convert to plotly and add accessibility
fig <- ggplotly(p, tooltip = "text")

# Add the keyboard accessibility in the plot
fig <- ro_ply_keyboard_nav_trendline(fig)
fig
```
### plotly Trendline
```{r plotly Trendline}
library(plotly)
 
fig <- plot_ly()
fig <- fig |>
  plotly::add_trace(
    data = data,
    x = ~Date_statistics,
    y = ~n,
    type = "scatter",
    mode = "markers",
    name = "Aantal cases",
    text = ~text,
    hoverinfo = "text"
  )
# Add line for moving average
fig <- fig |>
  plotly::add_trace(
    data = data,
    x = ~Date_statistics,
    y = ~moving_average,
    type = "scatter",
    mode = "lines",
    name = "7-daags gemiddelde",
    text = ~text,
    hoverinfo = "text"
  )

# Add the keyboard accessibility in the plot
fig <- ro_ply_keyboard_nav_trendline(fig)
fig
```

## Grouped Trendline
This section demonstrates how to make a grouped trendline created with ggplotly or plotly keyboard accessible. The `ro_ply_keyboard_nav_trendline()` function allows users to focus on individual points (not lines), with tooltips and ARIA labels for screen readers.

### Example data
```{r example data Grouped Trendline}
categories <- LETTERS[1:5]  # "A", "B", "C", "D", "E"
weeks <- seq(as.Date("2023-01-02"), as.Date("2023-05-29"), by = "week")

data <- tidyr::expand_grid(
  week = weeks,
  Category = categories
) |> 
  dplyr::mutate(
    n = dplyr::case_when(
      Category == "A" ~ round(100 + 10 * as.numeric(week - min(week)) / 7 + stats::rnorm(dplyr::n(), 0, 15)),
      Category == "B" ~ round(50 + 2 * as.numeric(week - min(week)) / 7 + stats::rnorm(dplyr::n(), 0, 7)),
      Category == "C" ~ round(80 + 5 * sin(dplyr::row_number() / 4) + stats::rnorm(dplyr::n(), 0, 10)),
      Category == "D" ~ round(200 - 5 * as.numeric(week - min(week)) / 7 + stats::rnorm(dplyr::n(), 0, 18)),
      Category == "E" ~ round(70 + 8 * cos(dplyr::row_number() / 5) + stats::rnorm(dplyr::n(), 0, 12))
    )
  )
data <- data |>
  dplyr::mutate(
    text = paste0(
      "Week: <b>", format(week, "%d-%m-%Y"), "</b><br>",
      "Categorie: <b>", Category, "</b><br>",
      "Aantal cases: <b>", format(n, big.mark = ".", decimal.mark = ",", scientific = FALSE), "</b>"
    )
  )
cat_names <- unique(data$Category)
```
### ggplotly Grouped Trendline
```{r ggplotly Grouped Trendline}
library(ggplot2)
library(plotly)
 
p <- ggplot(data, aes(x = week, y = n, color = Category, group = Category, text = text)) +
  geom_line() +
  geom_point() 

# Convert to plotly and add accessibility
fig <- ggplotly(p, tooltip = "text")

# Add the keyboard accessibility in the plot
fig <- ro_ply_keyboard_nav_trendline(fig)
fig
```

### plotly Grouped Trendline

```{r plotly Grouped Trendline}
library(plotly)
 


fig <- plot_ly()
for (cat in cat_names) {
  subset <- data[data$Category == cat, ]
  fig <- fig |>
    plotly::add_trace(
      data = subset,
      x = ~week,
      y = ~n,
      type = "scatter",
      mode = "lines+markers",
      name = cat,
      text = ~text,
      hoverinfo = "text"
    )
}

# Add the keyboard accessibility in the plot
fig <- ro_ply_keyboard_nav_trendline(fig)
fig
```

## Line Chart

This section demonstrates how to make a line chart created with ggplotly or plotly keyboard accessible. The `ro_ply_keyboard_nav_linechart()` adds keyboard focus to lines, and shows a focus-point.

### Example data
```{r example data Line Chart}
base_data <- tidyr::tibble(expand.grid(
  Jaar = 2001:2020,
  Geslacht = c("Man", "Vrouw")
)) |>
  dplyr::mutate(value = sample(1:50, dplyr::n(), TRUE))

total_data <- base_data |>
  dplyr::group_by(Jaar) |>
  dplyr::summarize(value = sum(value), .groups = "drop") |>
  dplyr::mutate(Geslacht = "Totaal")

data <- dplyr::bind_rows(base_data, total_data) |>
  dplyr::mutate(Geslacht = factor(Geslacht, levels = c("Man", "Vrouw", "Totaal")))

```
### ggplotly Line Chart
`ro_ply_keyboard_nav_linechart`
```{r ggplotly Line Chart}
library(ggplot2)
library(plotly)
 

p <- ggplot(data, aes(x = Jaar, y = value, color = Geslacht, group = Geslacht,
                      text = paste0(
                        "Geslacht: <b>", Geslacht,
                        "</b><br>Jaar: <b>", Jaar,
                        "</b><br>Aantal cases: <b>", value
                      ))) +
  geom_line() 

# Convert to plotly and add accessibility
fig <- ggplotly(p, tooltip = "text")

# Add the keyboard accessibility in the plot
fig <- ro_ply_keyboard_nav_linechart(fig)
fig
```
### plotly Line Chart
```{r plotly Line Chart}
library(plotly)
 

fig <- plot_ly()

for (geslacht in levels(data$Geslacht)) {
  subset <- data |> filter(Geslacht == geslacht)
  
  fig <- fig |>
    add_trace(
      data = subset,
      x = ~Jaar,
      y = ~value,
      type = "scatter",
      mode = "lines",
      name = geslacht,
      text = paste0(
        "Geslacht: <b>", subset$Geslacht,
        "</b><br>Jaar: <b>", subset$Jaar,
        "</b><br>Aantal cases: <b>", subset$value
      ),
      hoverinfo = "text"
    )
}

# Add the keyboard accessibility in the plot
fig <- ro_ply_keyboard_nav_linechart(fig)
fig
```

# Keyboard accessibility - ggplotly and plotly legends

By default, plotly legends are not keyboard accessible. 
The `ro_ply_keyboard_nav_legend()` function makes it possible to reach and toggle legend items using the keyboard, with visual focus indication and ARIA attributes for screen readers.

Note: For stacked bar graphs for ggplot, it is advised to disable the legend filtering (also with the mouse) altogether, as the image is not re-rendered after filtering, which can lead to 'floating' bars.

### Example data and plot (horizontal barchart from above)

The example data is used from the horizontal barchart as shown above.

```{r example data horizontal barchart for the legend}
data <- tidyr::tibble(
  expand.grid(`Age group` = c("A", "B", "C", "D", "E"), `Sex` = c("Man", "Vrouw")),
  `Number of Cases` = sample(1:50, 10, TRUE)
) 
library(ggplot2)
library(plotly)
 
p <- ggplot(data, aes(
  x = `Number of Cases`,
  y = `Age group`,
  fill = Sex,
  text = paste0(
    "Leeftijdsgroep: <b>", `Age group`, "</b><br>",
    "Geslacht: <b>", Sex, "</b><br>",
    "Aantal cases: <b>", `Number of Cases`, "</b>"
  )
)) +
  geom_col(position = "dodge") +
  labs(
    x = "Aantal cases",
    y = "Leeftijdsgroep",
    fill = "Geslacht"
  ) 

# Convert to plotly and add accessibility
fig <- ggplotly(p, tooltip = "text")

# Add the keyboard accessibility in the plot
fig <- ro_ply_keyboard_nav_barchart(fig)
```
After the figure is made, the legend can be made accessible.

```{r legend example horizontal barchart} 
# Add the keyboard accessibility to the legend
fig <- ro_ply_keyboard_nav_legend(fig)
fig

```


# ggplotly and plotly disable toolbar
The plotly modebar / toolbar can be distracting or inaccessible for keyboard users. 

The `ro_ply_disable_toolbar()` function removes the toolbar and disables legend popup further improving accessibility.

### Example data and plot 

The example data is used from the horizontal barchart as shown above.

```{r example data horizontal barchart for disabling the toolbar}
data <- tidyr::tibble(
  expand.grid(`Age group` = c("A", "B", "C", "D", "E"), `Sex` = c("Man", "Vrouw")),
  `Number of Cases` = sample(1:50, 10, TRUE)
) 
library(ggplot2)
library(plotly)
 
p <- ggplot(data, aes(
  x = `Number of Cases`,
  y = `Age group`,
  fill = Sex,
  text = paste0(
    "Leeftijdsgroep: <b>", `Age group`, "</b><br>",
    "Geslacht: <b>", Sex, "</b><br>",
    "Aantal cases: <b>", `Number of Cases`, "</b>"
  )
)) +
  geom_col(position = "dodge") +
  labs(
    x = "Aantal cases",
    y = "Leeftijdsgroep",
    fill = "Geslacht"
  ) 

# Convert to plotly and add accessibility
fig <- ggplotly(p, tooltip = "text")

# Add the keyboard accessibility in the plot
fig <- ro_ply_keyboard_nav_barchart(fig)
# Add keyboard accessibility to the legend
fig <- ro_ply_keyboard_nav_legend(fig)
```
Then the toolbar / modbar can be disabled, and the pop-up on filtering can be disabled as well, as this not accessible with the keyboard

```{r modbar example horizontal barchart} 
# Disable the modbar and filtering pop-up
fig <- ro_ply_disable_toolbar(fig)
fig
```
